VAIBHAV CHAWLA
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
library(dplyr)
library(ggplot2)
library(psych)
library(scatterplot3d)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv("wineQualityReds.csv",sep = ",")
```
#INTRODUCTION:
The given dataset has data of around 1599 red-wines samples with 
13 variables("X","fixed.acidity","volatile.acidity","citric.acid",
"residual.sugar","chlorides","free.sulfur.dioxide","total.sulfur.dioxide",
"density","pH","sulphates","alcohol","quality)
Summary of the dataset shows that "x" is a count variables and there are some 
intersting features to explore in the dataset like the quality of red-wine and 
may be after analysis, we can comment upon the factors that control the quality
of red-wine.
```{r echo=FALSE}
summary(df)
```


# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
qplot(x= as.factor(quality), data= df, geom = "bar")
```


Quality is an ordered factor variable most of the red whines in the data 
belong to quality 5 followed by quality 6. Higher quality red wines are less 
frequent as suggested by plot. 


```{r echo=FALSE}
qplot(x= alcohol, data= df, bins = 50)
```


Mostly alcohol content values lies b/w 9 and 10.5 and a peak, around at an 
alcohol content of 9.5. Also, there are some outliers towards higher alcohol 
content and a few at lower side. 

```{r echo=FALSE}
qplot(x= pH, data= df, bins = 100)
```


pH distribution is normal with very few outliers. 
Large chunk of red wines have pH values in the range (3.2-.3.3).
```{r echo=FALSE}
summary(df$pH)
subset(df,df$pH >4)

```


In the plot, there are some data points at extreme right. 
So, I checked them with a R-code to filter the data having ph>4. 
```{r echo=FALSE}
qplot(x= density, data= df, bins = 70)
```


Distribution of density is normal with few outliers.


```{r echo=FALSE}
qplot(x = volatile.acidity, data= df, bins= 80)
```

We need to use log scale in this case. 

```{r echo=FALSE}
qplot(x= volatile.acidity, data= df, bins= 80)+ scale_x_log10()
```


log(volatile acidity) has somewhat normal distribution.
A little cleaning of data will give a better plot.

```{r echo=FALSE}
qplot(x= fixed.acidity, data= df, bins = 70)
```

Here, we again need to use log transformation on our x-axis.

```{r echo=FALSE}
qplot(x= fixed.acidity, data= df, bins = 70)+scale_x_log10()
```


Distribution of fixed.acidity. 
I have used a log scale and it appears that the distribution is somewhat normal.

```{r echo=FALSE}
qplot(x= sqrt(residual.sugar), data= df, bins= 80,)
```


I have used a sqrt scale on x-axis. 
A large chunk of data is concentrated towards the left with a long tail
towards higher residual sugar levels. 

```{r echo=FALSE}
qplot(x= citric.acid, data= df, bins= 80,)
```

The distribution is skewed. Lets see the plot after axis transformation.

```{r echo=FALSE}
qplot(x= citric.acid, data= df, bins= 80,)+scale_x_sqrt()
```


Distribution is a bit bimodal in nature.
But, a lot of data values are absent in between.

```{r echo=FALSE}
qplot(x= sulphates, data= df, bins= 80)
```


Lets use log scale on x-axis.

```{r echo=FALSE}
qplot(x= sulphates, data= df, bins= 80)+scale_x_log10()
```


Again data is somewhat normally distributed with few ouliers that can be
removed to get a better plot.
```{r echo=FALSE}
qplot(x= total.sulfur.dioxide, data= df, bins= 80)
```


The data is highly skewed towards left. We need to use log scale.

```{r echo=FALSE}
qplot(x= total.sulfur.dioxide, data= df, bins= 80)+ scale_x_log10()
```


I have used log scale to even the distribution which was previously 
skewed towards left.

```{r echo=FALSE}
qplot(x= chlorides, data= df, bins= 80)
```

Graph without log scale


```{r echo=FALSE}
qplot(x= chlorides, data= df, bins= 80)+scale_x_log10()
```


I have plotted chloride content on a log-scale.
Distribution is normal with some outliers scattered around higher
chloride content.

# Univariate Analysis

### What is the structure of your dataset?
The dataset(1599X13) contains information of 1599 red-wines with 13 variables
("X","fixed.acidity","volatile.acidity","citric.acid","residual.sugar",
"chlorides","free.sulfur.dioxide","total.sulfur.dioxide","density","pH",
"sulphates","alcohol","quality).
"x" is a count variable.
"Quality is an ordered factor variable with lowest quality being 3 and
heighest being 8.           
            
### What is/are the main feature(s) of interest in your dataset?
I would really like to work out the relationahip between quality and alcohol
content of red wines.
Does higher quality red wines have higher alcohol content or not. 
Also the variation of pH with alcohol content. 

### What other features in the dataset do you think will help support your 
###investigation into your feature(s) of interest?
What holds the control over pH of red wines i.e. free.sulhur.dioxide, 
volatile.acidity, sulphates or alcoho.
Also, variation of parameters in quality 5 and 6(most counts). 
What features have strong co-relations.
And, variation of density with quality.

### Did you create any new variables from existing variables in the dataset?
No, I didn't create any new variables from existing variables in the dataset. 
But, I converted quality into 3 groups i.e. 3-5, 5-7,7-9.
And, grouped parameters by quality and then summarised parameters in the 
quality group (5-7). 

### Of the features you investigated, were there any unusual distributions? 
###Did you perform any operations on the data to tidy, adjust, or change the
###form of the data? If so, why did you do this?
I used log scale and sqrt scale in univariate plots to obtain somewhat
normal distribution.


# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x= as.factor(quality), y=pH), data= df)+geom_boxplot()
```


It is evident from the plot that pH varies alot with quality 
and there is no strong relation indicated by the plot.
Also, thereare some outliers present in the data.
Below is the correlation test by pearson method.

```{r echo=FALSE}
cor.test(df$pH,df$quality, method = "pearson")
```
 There is no strong correlation b/w pH and quality.
 Lets summarise pH variation with quality of alcohol.
```{r echo=FALSE}
describeBy(df$pH, group= df$quality, mat= TRUE)
``` 
 
 
 The summary stats show that the mean pH varies b/w 3.26-3.40. 
 Also, mean pH is decreasing with increase in quality of red wines. 
 
```{r echo=FALSE}
ggplot(aes(x= quality, y= alcohol), data= df)+geom_jitter(alpha= 0.6)+
  geom_line(stat= "summary", fun.y =mean, linetype =2, color = 'blue')+
  geom_line(stat= "summary", fun.y = quantile, fun.args = list(probs = .9)
            ,color= "blue")
```


Plot between alcohol content and quality.
Though for each quality of red-wine there is a fair variation of alcohol content
but,the plot suggests that mean alcohol content is increasing with quality. 
The alocohol content dips from quality 4 to quality 5.
Lets check how much they are correlated !

```{r echo=FALSE}
cor.test(df$alcohol,df$quality, method = "pearson")
```


Pearson's r is 0.47>0.3. 
So, we can say that, there is moderately strong coorelation b/w alcohol content
and quality of alcohol.
Lets summarise alcohol content with quality of red wine.
```{r echo=FALSE}
describeBy(df$alcohol, group= df$quality, mat= TRUE)
``` 


As, suggested by the correlation plot, mean alcohol content increases with 
increase in quality of red wine. 
```{r echo=FALSE}
ggplot(aes(x= quality, y= density), data= df)+geom_jitter(alpha= 0.55)+
  geom_line(stat= "summary", fun.y =mean, linetype =2, color = 'blue')+
  geom_line(stat= "summary", fun.y = quantile, fun.args = list(probs = .9),
            color= "blue")
```


Density is decreasing initially but then it increases with increasing quality 
for a while only to fall again with quality
From the plot, it does not appears that the two parameters are much correlated.  
Lets check it statistically!
```{r echo=FALSE}
cor.test(df$density,df$quality, method = "pearson")
```


 correlation is not quite relevant.A weak negative correlation isn't much of
 interest.
 Lets summarise density distribution with quality.
 
```{r echo=FALSE}
describeBy(df$density, group= df$quality, mat= TRUE)
``` 


Mean density varies from .995 to .997.

```{r echo=FALSE}
ggplot(aes(x= quality, y= volatile.acidity), data= df)+geom_jitter(alpha= 0.55)+
  geom_line(stat= "summary", fun.y =mean, linetype =2, color = 'blue')+
  geom_line(stat= "summary", fun.y = quantile, fun.args = list(probs = .9),
            color= "blue")
```


It appears from the plot that, as quality of red wine increases, 
volatile acidity decreases. 
Lets check the pearson's r.

```{r echo=FALSE}
cor.test(df$volatile.acidity,df$quality, method = "pearson")
```
Yes, a negative moderately strong correlation b/w the two parameters is observed.
Lets summarise this variation.
```{r echo=FALSE}
describeBy(df$volatile.acidity, group= df$quality, mat= TRUE)
```


Mean value of volatile acidity varies b/w 0.40-0.89.
Standard deviation is quite small too.

```{r echo=FALSE}
ggplot(aes(x= alcohol, y= pH), data= df)+geom_point()
```


Distribution of pH vs alcohol is widely scattred.
plot suggests that alcohol content doesn't control pH of red wines.
Quite interesting !

```{r echo=FALSE}
cor.test(df$pH,df$alcohol, method = "pearson")
```


As evident from the plot and  from a very small value of pearson's "r", 
correlation is very weak.

```{r echo=FALSE}

ggplot(aes(x= alcohol, y= density), data= df)+geom_point()
```


It appears that there is a negative relation b/w the two parameters.
However, data is quite scattered.
Lets check our assumption!
```{r echo=FALSE}
cor.test(df$density,df$alcohol, method = "pearson")
```

Wow! A negative correlation.

```{r echo=FALSE}
ggplot(aes(x= pH, y= fixed.acidity), data= df)+geom_point()+
geom_line(stat= "summary", fun.y =mean, linetype =2, color = 'blue')
```


The plot suggests strongly negative correlation b/w pH and fixed acidity. 
Lets confirm it with pearson's r !
```{r echo=FALSE}
cor.test(df$fixed.acidity,df$pH, method = "pearson")
```


Strong ngative co-relation b/w the two parameters.

```{r echo=FALSE}
ggplot(aes(y= pH, x= citric.acid), data= df)+geom_point()
```


Lets check the correlation.

```{r echo=FALSE}
cor.test(df$pH,df$citric.acid, method = "pearson")
```


A moderately strong but negative correlation.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
###investigation. How did the feature(s) of interest vary with other features 
###in the dataset?
 I was interested in studying the relationship b/w alcohol content and 
 quality of red wines.
 From the plot as well as from pearson's r, the two parameters have quite 
 strong correlation b/w them.
 Mean alcohol content increases with the quality of red wines. 
 Quality isn't strongly related to pH or density but related with 
 volatile acidity.
 Alcohol content varied widely with pH but negatively related with density.
 
### Did you observe any interesting relationships between the other features
###(not the main feature(s) of interest)?
Also, before analysis I thought that alcohol content and pH would be correlated
but surprisingly fixed.acidity controls the pH.
The relationship between fixed.acidity and pH is quite strong too.

### What was the strongest relationship you found?
Fixed.acidity and pH

# Multivariate Plots Section
```{r}
breaks <- c( 3,4,6,9 )
df$quality.group <-cut(df$quality,breaks)
quality_group<-group_by(df,quality.group)
df.group <-summarise(quality_group,mean_alcohol =mean(alcohol),mean_pH= mean(pH),
                     mean_fixed.acidity= mean(fixed.acidity),
                     mean_volatile.acidity= mean(volatile.acidity),
                     mean_density= mean(density),n= n())
head(df.group)
```


I divided the data according to quality to get the mean values of parameters
that belong to various quality ranges.In the interset group variation of 
parameters is as follows 
Mean pH = 3.311296
mean alcohol = 10.252
mean fixed acidity = 8.254
Mean volatile acidity = 0.5385
Mean Density =0.99686
NA value group isn't useful for us.

```{r echo=FALSE}
shapes = c(15,16,17) 
shapes<- shapes[as.numeric(df$quality.group)]
colorcb = c("red","cyan","green")
colorcb<-colorcb[as.numeric(df$quality.group)]
with(df, {
  scatterplot3d(density,alcohol,volatile.acidity, pch = shapes,
                color = colorcb,angle= 65)
  })
```


I have created a 3D plot between density, volatile acidity and alcohol content.
I have used the quality group created above to colour code the data according
to quality.

```{r echo=FALSE}
ggplot(data =df, aes(x=volatile.acidity, y=sulphates, size = alcohol, 
                color=factor(df$quality))) + geom_point()
```


It looks like the higher quality red wines tend to be concentrated in the top
left of the plot. This tends to be where the higher alcohol content 
(larger dots) are concentrated as well.

```{r echo= FALSE}
ggplot(data = df , aes(x = sulphates, y = alcohol,colour=factor(df$quality))) + 
  geom_density2d(bins=2) + scale_color_brewer() + 
  geom_point(color="black", alpha=0.1) 
```


Summarising quality using a contour plot of alcohol and sulphate content.
Higher quality red wines are generally located near the upper right of the
scatter plot (darker contour lines) wheras lower quality red wines are 
generally located in the bottom right.


```{r echo=FALSE}
ggplot(aes(x= alcohol, y= density), data= df)+
geom_line(aes(color= quality.group),stat= "summary", fun.y =median)
```


I have plotted alcohol content and density with quality to see the
relation b/w the three parameters.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of 
###the investigation. Were there features that strengthened each other
###in terms of looking at your feature(s) of interest?

The  3D scatter plot and lineplot shows the relation b/w 2 or more parameters as 
a function of quality of red wines.

Summarising quality using a contour plot of alcohol and sulphate content.
Higher quality red wines are generally located near the upper right of the
scatter plot (darker contour lines) wheras lower quality red wines are 
generally located in the bottom right.

### Average value of parameters with quality.
Average value of parameters in the most frequent 
quality range(most of the red wines belong to this group):
Mean pH = 3.311
mean alcohol = 10.252
mean fixed acidity = 8.25
Mean volatile acidity = 0.53
Mean Density =0.99686

### Were there any interesting or surprising interactions between features?
Not really. 

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths
###and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
qplot(x= as.factor(quality), data = df, geom= 'bar',
      main = "Quality of Red-wines",  
  	xlab="Quality", ylab="Numberof red-wines")
```

### Description One
The first plot shows the quality of red wines. Most red wines have a
quality of 5 or 6.
Very low quality and high quality red-wines are less frequent.
```{r echo=FALSE}
summary(df$quality)
```
Mean quality is 5.636 while median is 6.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x= quality, y= alcohol), data= df)+
  geom_jitter(alpha= 0.6)+
  geom_line(stat= "summary", fun.y =mean, linetype =2, color = 'blue')+
  geom_line(stat= "summary", fun.y = quantile, fun.args = list(probs = .9),
            color= "blue")+
  labs(title="Distribution of alcohol content with quality", 
       x="quality", 
       y="alcohol (% by volume)")
  
```

### Description Two
The plot b/w alcohol content and quality of red-wines.The plot shows a quite 
strong co-relation b/w the two parameters. the mean alcohol content 
shows the increasing nature of alcohol content with quality of red-wines.


```{r echo=FALSE}
describeBy(df$alcohol,group = df$quality, mat = TRUE)
```
Summary of variation of alcohol content with quality.



### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(data =df, aes(x=volatile.acidity, y=sulphates, size = alcohol, 
                color=factor(df$quality))) + geom_point()+
labs(title="Relationship between sulphates,volatile.acidity, alcohol & quality", 
       x="volatile acidity (acetic acid - g / dm^3)", 
       y="sulphates (potassium sulphate - g / dm3)")
```

### Description Three
The plotshowsthe relationship between sulphates, volatile.acidity,
alcohol and quality. The alcohol content is encoded as size and quality as color 
 the higher quality red wines tend to be concentrated in the top left of 
 the plot. This tends to be where the higher alcohol content (larger dots) 
 are concentrated as well.

```{r, echo=FALSE}
head(df.group)
```

variation of parameters within quality groups.

# Reflection
The dataset contains information of about 1599 redwines with 13 variables to 
represent its characteristics.
Mostly parameters followed somewhat normal distribution.
Mostly redwines in the data belong to quality 5& 6.
I studied the relationship b/w quality of red wines and its various parameters 
and relationship b/w various parameters.
Lastly, I summarised the data into three quality groups.
The quality of redwines was correlated with alcohol content and volatile acidity.
Also, I assumed that pH would be controlled by alcohol content,
but to my surprise pH was correalted to citric acid and fixed acidity.
Not to forget, alcohol content and density were related too.

I struggled with multivariate analysis. It was bit difficult to use three-four
variables to use at the same time to draw plots. But, a little research here 
and there worked finally.Also, it being my very first experience in R, it was
a great experience exploring functions.

Using the insights from correlation coefficients provided by the paired plots,
it was interesting exploring quality using density plots with a different color
for each quality. Once I had this plotted it was interesting to build up the
multivariate scatter plots to visualize the relationship of different variables
with quality by also varying the point size, using density plots on the x and y
axis, and also using density plots.

A next step would be to develop a statistical model to predict red wine
quality based on the data in this dataset.
